### 2017-11-21 paper/response_to_reviewers.tex

The changes have also been made throughout the manuscript as generally suggested by all four reviewers.

\item{\textcolor{blue}{My main concern is that what disparity actually means is left to the reader’s imagination – it is not defined or discussed anywhere.
Disparity simply means lack of similarity or equality.
While the author is probably considering the dispersion of points in a multidimensional space as a measure of dissimilarity, it is important to realise the ambiguity of the word and more explanation is needed early in the introduction (and foremost in the abstract).
Specifically, the lines 58-61 need to be revised, as they simply don’t convey to the reader what is being presented here as disparity.
Say what “specific aspect of it [multidimensional space]” you are referring to.
Also, the different definitions of disparity alluded to on lines 61-63 should be summarised.
Furthermore, the manuscript is lacking a table summarising the metrics of disparity that this package can work with (which I suspect it was supposed to have).}}
\label{define_disparity}

%TODO improve disparity definition in the abstract

As reviewer 4 (major comment \ref{nebulous}) also underlines: ``disparity is a broad, nebulous concept''.
This observation was the genesis for this package and I am currently working on a collaborative review to define the borders (or the absence thereof) of this concept.
I feel that explaining this would be out of the scope of this software description paper.
However, I have tried to give specific examples each time I refer to disparity as a way to measure these ``specific aspect of the space'' and I have also added a clearer definition of what I mean by disparity throughout this manuscript.

@@

I've reflected these changes in the manual with a subsection in the chapter ``Specific tutorials'' and in manuscript as follows:

\textit{Note that the input matrix is not restricted to an ordinated matrix but can be any kind of matrix as long as its rows represent elements (e.g. the multidimensional space can be a distance matrixl; Close et al., 2015).} @@@

\item{\textcolor{blue}{Picking up on the first point from above, there are also additional post-ordination disparity metrics that have been used in the literature but are not available here. For example, hypercube volume, minimum spanning tree length etc. I would also urge the author to cite the work of Mike Foote, who made some early important contributions to disparity in general, and has written papers that discuss some of these potential metrics (e.g., Foote 1997, Annual Review of Ecology and Systematics, 28, 129–). I believe Wills (2001, cited) also discusses a long list of metrics.}}

I agree, there are many more metrics commonly used for disparity analysis not implemented in this package.
This was a conscious choice: due to the really high number of metrics available (especially when combining different ones), I designed the package to be modular hence allowing users to implement their own metrics (see response to reviewer 1 minor comment \ref{table_missing}).
I have precise this using the hypercube volume example in the manuscript:

@@

\item{\textcolor{blue}{L296 - Should be "increase".}}

I've fixed this typo.

\item{\textcolor{blue}{L357 - Should be "for such analyses".}}

I've removed this last sentence (see reviewer 1 minor comment \ref{remove_last_sentence}).

@@

I have worked towards making the manuscript more useful for non-morphologists.
I have generalised the input description as suggest by this reviewer, minor comment \ref{rows}; added a glossary table and a brief example using ecological data.
%TODO: add Diaz's example?
% %TG: adding an ecology example (removing manual and vignette description)

\item{\textcolor{blue}{The referencing is poor; why cite Close et al for the concept of pairwise distances?
I'm pretty sure PCO existed before Brusatte et al, and PCA certainly before 2012, etc.
Throughout I was frustrated with this; the referencing ignores a century of work outside this specific field.
I appreciate there merit in citing relevant literature for your field, but that doesn't absolve one of a duty to accurately attribute ideas to those that came up with them.
If you are unfamiliar with developments in ideas beyond your particular field, I would suggest Legendre \& Legendre's *Numerical Ecology* textbook as a good source for original work relevant to what is presented here.}}

I thank this reviewer for sharing the excellent Legendre \& Legendre reference (that was definitely a reference I was missing) and I've accurately credited the statistical methods throughout the manuscript - see my response to reviewer 2, minor suggestion \ref{PCO_cites}.

@@

\item{\textcolor{blue}{The comparison or critique of packages that, in the eyes of the author, restrict users grated a little with me. As a developer of one of the packages named, I feel that particular critique is unwarranted for several reasons;
\begin{itemize}
\item i) **vegan** doesn't restrict you at all, as with anything in R you are free to pick and choose ordination methods, dissimilarity measures etc and then do what you want with the scores,
\item ii) **vegan** doesn't provide disparity measures because it was never intended for such usage, but it is general enough that a user could do whatever they want,
\item and iii) like any software, the **dispaRity** author's software included, you have to provide defaults and **vegan**'s defaults reflect the intended audience of community ecology; that's a feature, as it is of **dispaRity**. As such the argument set up on page 6 is some straw man and unnecessary to motivate your package.
\end{itemize}
}}

I apologise for the unfounded critique and I have toned down this sentence (see reviewer 1, minor suggestion \ref{tone_down}).
% Additionally, I have added more functionalities using vegan and geiger %TODO!
% %TG: Tone down the critics to other packages

@@

\item{\textcolor{blue}{You don't really define disparity anywhere, except in very loose terms.
This needs making more explicit; I appreciate that disparity is a broad, nebulous concept, but you need to help the reader that may not be familiar with your subject area.
Either that or as mentioned above, refocus the paper on those readers already versed in this particular type of analysis.}}
\label{nebulous}

I have tried to define disparity with more examples and references to be more familiar to a broader range of readers (See reviewer 1, major suggestion \ref{define_disparity}).

@@

\item{\textcolor{blue}{L116 There is either a missing reference or a missing table, I presume the latter as no existing table matches the textual description in and around this line.}}

Sincerest apologies! See reviewer 1 minor comment \ref{table_missing}.

\item{\textcolor{blue}{L134 Needs "an" between "as" and "argument". L210 delete "sheer". L213 "longer" not "bigger" and "altitude" not "altitudes". L218 "longer" not "bigger"}}

I've fixed these typos (see also reviewer 2 minor comment \ref{altitude}).

@@

\item{\textcolor{blue}{L252 and throughout the code: insert straight quotes. There is a good chance these curly quotes will get replaced with some non-ASCII character and be impossible to paste into R.}}

I've converted all quotes in quoted code lines as ASCII straight quotes.

\item{\textcolor{blue}{In the example around L236--249, can you justify the metric used here? Is the sum of variances along components of an ordination a valid measure of the dispersion in multivariate space of samples? You don't, or I didn't see, reference to original work that justifies this, and it's not something I have come across despite implementing some approaches for testing multivariate dispersion in the **vegan** package. Doesn't this suffer the same problem as confidence bars in an x-y scatterplot where the bars have been defined with reference to only one axis at a time? The solution is a confidence ellipse, which is often quite different to the interval implied by separate confidence interval bars for each axis.}}

This metric (the sum of each ordination axis variance) is commonly used in palaeobiology (since Foote 1991 and Wills et al. 1994) probably due to a ``citation diffusion effect'' (i.e. future authors replicating convincing methods) and partly due to its pretty good behaviour with palaeontological data (e.g. missing data, Ciampaglio 2001).
However, as this reviewer high

@@

\item{\textcolor{blue}{L326 Where do the 300 samples come from as implied by the residual degrees of freedom. Table 3 suggests 15 + 9 + 13 = 37 samples in this analysis. I'm clearly missing something here. Also, this is an omnibus test, which if we believe the printed values (see previous re DF) only says there is some effect of `subsamples`. Can this analysis be extended to point estimates for mean variance (again assuming you can justify this sum of individual variances method) and post-hoc comparisons of pairwise `subsample` differences, which would be quite useful?}}

%TODO correct/explain this 300 subsamples + add a posthoc t-test.

\item{\textcolor{blue}{If you need to save space to add an ecological example, cut the *Manuals and vignettes* section and also the *Data simulations* section. The later is somewhat tangential to the main thrust of the paper and you cover this in the vignette/bookdown resources.}}
\label{remove_simulations}

I have removed the simulations part as suggested by this reviewer.
However, following the encouraging suggestions of reviewer 2 (major suggestion \ref{simulations}) I have developed in much more details the simulation algorithms in the package manual.

### 2017-11-21 paper/dispRity_paper_highlighted.tex
response update

In this example, the multidimensional space is defined as a morphospace: the ordination of the distances among discrete morphological characters for 50 mammal species \citep[from][]{beckancient2014}.
Additionally, we can define disparity as the sum of the variances on each dimension \citep{Wills1994} that will represent an aspect of the the volume of the morphospace (i.e. a large sum of variances approximates a large morphospace and vice versa).

>>>>

In this example, the multidimensional space is defined as a morphospace: the ordination of the distances among discrete morphological characters for 50 mammal species \citep[from][]{beckancient2014}.
Additionally, we can define disparity as the sum of the variances on each dimension \citep{foote1991morphological,Wills1994} that will represent an aspect of the the volume of the morphospace (i.e. a large sum of variances approximates a large morphospace and vice versa).
